---
title: "The RMST BART package"
author: 
  - Nicholas C. Henderson
output: 
  pdf_document:
    fig_caption: yes
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rmstbart)
library(AFTrees)
```

## Overview

`rmstbart` is an R package implementing an approach for using Bayesian additive regression trees (BART) to 
conduct inference on the restricted mean survival time (RMST). 


To use `rmstbart`, one needs survival outcome data and covariates associated with each
survival outcome. We will use the following notation for the survival data of interest:

* $n$ - total sample size

* $T_{i}$ time-to-failure for an event of interest for individual $i$ (not always observed).

* $C_{i}$ time of right-censoring for individual $i$ (not always observed). We cannot observe both $T_{i}$ and $C_{i}$,

* $U_{i} = \min\{ T_{i}, C_{i} \}$ the **observed** follow-up time for individual $i$. 

* $\delta_{i} = I(T_{i} \leq C_{i})$ the **observed** event indicator for individual $i$. 

* $\mathbf{U} = (U_{1}, \ldots, U_{n})$ the vector of follow-up times for all individuals 

* $\boldsymbol{\delta} = (\delta_{1}, \ldots, \delta_{n})$ the vector of event indicators for all individuals. 

* $\mathbf{x}_{i} = (x_{i1}, \ldots, x_{ip})$ the vector of baseline covariates associated with the survival outcome $(U_{i}, \delta_{i})$.


Of primary interest are the **restricted mean survival times (RMST)** for survival times $T_{i}$.
For an individual with covariate vector $\mathbf{x}_{i}$, the RMST $\mu_{\tau}(\mathbf{x}_{i})$ with restriction point $\tau$ for the survival time $T_{i}$ is defined as the expectation of the minimum of $T_{i}$ and $\tau$ conditional on the value of the covariate vector $\mathbf{x}_{i}$. Specifically,
the RMST function of interest is
\begin{equation}
\mu_{\tau}( \mathbf{x} ) = E[ \min\{ T_{i}, \tau \} \mid \mathbf{x}_{i} = \mathbf{x} ] 
\end{equation}

---

The idea behind `rmstbart` is to let the following
inverse probability of censoring weighted (IPCW) loss function
drive the inference for the unknown RMST function $\mu_{\tau}(\mathbf{x})$:
\begin{eqnarray}
\eta \sum_{i=1}^{n} \frac{\delta_{i}}{G(U_{i}^{\tau}|\mathbf{x}_{i})}\big[ U_{i}^{\tau} - \mu_{\tau}(\mathbf{x}_{i}) \big]^{2} \nonumber
\end{eqnarray}


* $G(t | \mathbf{x}_{i}) = P( C_{i} > t | \mathbf{x}_{i})$ denotes the survival distribution of the censoring time, conditional on $\mathbf{x}_{i}$.


## rmstbart Usage

You do need to install the `AFTrees` package before using `rmstbart`. This can be done using
the `devtools` package:
```{r, eval=FALSE}
library(devtools)
install_github("nchenderson/AFTrees")
```

### The METABRIC Data

To illustrate the use of the package, we will use the `METABRIC` dataset, which is available within
the `rmstbart` package. To load this dataset, just use the following command:
```{r}
data(METABRIC)
```
This dataset has 1839 observations and 32 variables
```{r}
dim(METABRIC)
```
The survival outcomes are stored in the variables `overall_survival_months` (the follow up times)
and `overall_survival` (the event indicators)
```{r}
METABRIC$overall_survival_months[1:10]
METABRIC$overall_survival[1:10]
```

### Format of rmstbart function

The standard format required for using the `rmstbart` function is the following
```{r, eval=FALSE}
rmstbart(times, delta, x.train)
```

* `times` - vector containing follow-up times

* `delta` - vector containing event indicators

* `x.train` - model matrix (the ith row of this matrix is $\mathbf{x}_{i}$)

As an example, let's create a model matrix assuming
that we want to use the following four variables
from METABRIC: `tumor_stage`, `tumor_size`, `age_at_diagnosis`, and `TP53`.
One way to do this is with the following code:
```{r}
X1 <- model.matrix(~ tumor_stage + tumor_size + age_at_diagnosis + TP53 -1, data=METABRIC)
```

* Note that you we used the `Surv` formatting from the `survival` package to
handle the survival outcomes.

* Note also that we used -1 to remove the intercept from `X1`. There is 
no need to have intercept terms when using `rmstbart`. They will not be used.

The matrix `X1` has 1839 rows and 8 columns (`tumor_stage` has 5 possible levels),
```{r}
dim(X1)
```
and the first 5 rows of this matrix look like the following
```{r}
head(X1)
```

Note that you are not required to enter a value of $\tau$ in order to run \verb"rmstbart". 
If no value of $\tau$ is provided, \verb"rmstbart" uses the maximum observed survival
time (i.e., the maximum value of $U_{i}$ among observations with $\delta_{i}=1$) as the value of $\tau$.
If you would like to use a particular value of $\tau$, for example $\tau = 300$, you would
call `rmstbart` using the following format
```{r, eval=FALSE}
rmstbart(times, delta, x.train, tau=300)
```

### Output from rmstbart

Now, to estimate the RMST function with these four covariates, we can use the following code
```{r}
bart_obj1 <- rmstbart(times=METABRIC$overall_survival_months, 
                      delta=METABRIC$overall_survival,
                      x.train=X1,
                      tau=300,
                      printevery=10000)
## pick large value of printevery so it does not print out
## frequent MCMC updates
```
`rmst_bart` returns an object of class "bart\_rmst". 
This is essentially a list with a number of components: 
```{r, eval=FALSE}
names(bart_obj1)
```
To start, the main components of `bart_obj1` that will typically be of interest are `yhat.train` and
`yhat.train.mean`. 

* `yhat.train` - This is a `ndpost` $\times n$ matrix. Here, `ndpost` is the number of posterior draws (the default value of `ndpost` in `rmstbart` is 1000)
$$
yhat.train = 
\begin{bmatrix}
\mu_{\tau}^{(1)}(\mathbf{x}_{1}) & \mu_{\tau}^{(1)}(\mathbf{x}_{2}) & \ldots & \mu_{\tau}^{(1)}(\mathbf{x}_{n}) \\
\mu_{\tau}^{(2)}(\mathbf{x}_{1}) & \mu_{\tau}^{(2)}(\mathbf{x}_{2}) & \ldots & \mu_{\tau}^{(2)}(\mathbf{x}_{n}) \\
\vdots & \vdots & \ddots & \vdots \\
\mu_{\tau}^{(S)}(\mathbf{x}_{1}) & \mu_{\tau}^{(S)}(\mathbf{x}_{2}) & \ldots & \mu_{\tau}^{(S)}(\mathbf{x}_{n})
\end{bmatrix}
$$
In the `yhat.train` matrix, $\mu_{\tau}^{(s)}(\mathbf{x}_{i})$ is the $s^{th}$ draw from the posterior distribution of
the RMST function evaluated at $\mathbf{x}_{i}$: $\mu_{\tau}(\mathbf{x}_{i})$.

* `yhat.train.mean` - This is a $n \times 1$ vector containin the posterior means of $\mu_{\tau}(\mathbf{x}_{i})$:
$$
yhat.train.mean = 
\begin{bmatrix}
\frac{1}{S}\sum_{s=1}^{S} \mu_{\tau}^{(s)}(\mathbf{x}_{1}) \\
\frac{1}{S}\sum_{s=1}^{S} \mu_{\tau}^{(s)}(\mathbf{x}_{2}) \\
\vdots  \\
\frac{1}{S}\sum_{s=1}^{S} \mu_{\tau}^{(s)}(\mathbf{x}_{n}) 
\end{bmatrix}
$$

* We can name the vector posterior means as `mu_hat`
```{r}
mu_hat <- bart_obj1$yhat.train.mean
mu_hat[1:5]
```
For example, the posterior mean for the RMST for an individual with the following covariates: a stage 2 tumor, a tumor size of 22.0, an age of 76.6 years,
and a TP53 expression level of 0.35 is about $120.7$ months. 


* We can obtain the vector of posterior credible intervals by using the `CredibleIntervals` function:
```{r}
mu_credibles <- CredibleIntervals(bart_obj1)
```
This returns an $n \times 2$ matrix with $95\%$ credible intervals for $\mu_{\tau}(\mathbf{x}_{1}), \mu_{\tau}(\mathbf{x}_{2}), ...$
```{r}
head(mu_credibles)
```

---

* You can **plot** all of the RMST posterior means and associated credible intervals by 
directly inputting the `bart_obj1` object into the `plot` method
```{r, fig.cap="Plotting all RMST posterior means and associated credible intervals"}
plot(bart_obj1, xlab="RMST in Months")
```

### Variable Importance

With BART, there are $H$ trees (the default is $H=200$) involved in "fitting" the RMST function within each MCMC iteration.
To explore variable importance, it is common to look at the number of times a variable is used
within a tree (for each MCMC iteration). Variables that are frequently used as tree splitting variables
are generally thought to be more "important".

These **variable counts** are contained in the `varcount` of the returned `bart_rmst` object:
```{r}
dim(bart_obj1$varcount)
```
`bart_obj1$varcount` will be a `ndpost` $\times p$ matrix. The $k^{th}$ row of this matrix contains
the number of trees (out of 200) each variable was used in, within the $k^{th}$ MCMC iteration:
```{r}
head(bart_obj1$varcount)
```
To summarize this, you can look at the posterior mean of the variable counts
```{r}
colMeans(bart_obj1$varcount)
```

```{r, echo=FALSE, fig.cap="Average variable counts across trees."}
tree_counts <- colMeans(bart_obj1$varcount)
par(mar=c(8, 4, 4, 2))  # Increase bottom margin to fit long labels
barplot(height=tree_counts, names.arg=names(data), las=2, col="blue", cex.names=0.8)
par(mar=c(4.1, 4.1, 2.0, 1.0))
```


### Censoring Weights

`rmstbart` also returns the matrix of censoring weights. This will be
a matrix with dimensions `ndpost` $\times n_{e}$, where $n_{e}$ 
is the number of observed events. This is contained in the `censoring.weights`
component of `bart_obj1`



### Looking at the RMST function at test points

In the previous example of running `rmst_bart`, it returns
posterior draws $\mu_{\tau}^{(s)}(\mathbf{x})$ of the RMST function 
only at the observed covariate values $\mathbf{x}_{i}$. If 
you want posterior draws for $\mu_{\tau}(\mathbf{x})$ at other
(non-observed) locations of $\mathbf{x}$, you will need 
to add a "test" model matrix as one of the arguments in \verb"rmst_bart". 
For example, suppose we wanted to look at the RMST function
for the following two additional sets of covariate values:

* `tumor_stage=3`, `tumor_size=50.3`, `age_at_diagnosis=75`, `TP53=1`

* `tumor_stage=2`, `tumor_size=10.3`, `age_at_diagnosis=55`, `TP53=0`

The way to do this is to first define the following $2 \times 8$ test model matrix of
interest
```{r}
newdf <- data.frame(tumor_stage = factor(c("3", "2"), levels=c("0", "1", "2", "3", "4")), 
                    tumor_size = c(50.3, 10.3),
                    age_at_diagnosis = c(75, 55), 
                    TP53 = c(1, 0))

X1test <- model.matrix(~ tumor_stage + tumor_size + age_at_diagnosis + TP53 -1, 
                       data=newdf)
X1test
```

* Then, just set the `x.test` argument equal to `X1test` when calling `rmstbart`. You
can otherwise keep all the arguments the same:
```{r}
bart_obj_test <- rmstbart(times=METABRIC$overall_survival_months, 
                          delta=METABRIC$overall_survival,
                          x.train=X1,
                          x.test=X1test,
                          tau=300,
                          printevery=10000)
```

* The components of `bart_obj_test` that you will want to look at are `yhat.test` and `yhat.test.mean`.
These have the same interpretation as `yhat.train` and `yhat.train.mean` except the RMST function 
is now being evaluated on the test points
```{r}
dim(bart_obj_test$yhat.test)
```

Posterior means of the RMST function for these two test covariate values:
```{r}
bart_obj_test$yhat.test.mean
```
$95\%$ credible intervals for the RMST function at these two test covariate values
```{r}
upper_cred <- apply(bart_obj_test$yhat.test, 2, function(x) quantile(x, probs=0.975))
lower_cred <- apply(bart_obj_test$yhat.test, 2, function(x) quantile(x, probs=0.025))
cbind(lower_cred, upper_cred)
```











